---
source: Rmd
title: "Summary and Exploration"
teaching: 15 
exercises: 50
questions:
- "TODO"
objectives:
- "TODO"
keypoints:
- "TODO"
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("05-")
```

## Summary

TODO: 

## Exploration

TODO: 

## Toxicity Prediction

[paper](https://doi.org/10.1021/ci034032s)

```{r, eval = FALSE}
download.file("https://www.openml.org/data/get_csv/52534/topo_2_1.arff", here("data", "topo21.csv"))
```

```{r, eval = FALSE}
library(tidyverse)
library(here)
topoInfo <- read_csv(here("data", "topo21.csv")) %>%
  rename(toxicity = oz267) %>% # toxicity is what we want to predict
  filter(toxicity > 0.1) # remove outlier
```

## Challenge: Decision forest, and evaluate RMSE

```{r, eval = FALSE}
trainSize <- round(0.80 * nrow(topoInfo))
set.seed(1234) 
trainIndex <- sample(nrow(topoInfo), trainSize)
trainDF <- topoInfo %>% dplyr::slice(trainIndex)
testDF <- topoInfo %>% dplyr::slice(-trainIndex)
```



```{r, eval = FALSE}
library(randomForest)
set.seed(4567)
toxfor <- randomForest(toxicity ~ ., data = trainDF, mtry = 10) # retrict mtry!
print(toxfor)
```




```{r, eval = FALSE}
pToxicity <- predict(toxfor, testDF)
rfErrors <- pToxicity - testDF$toxicity
tibble(`Predicted Toxicity` = pToxicity, Error = rfErrors) %>%
  ggplot(aes(x = `Predicted Toxicity`, y = Error))  +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 0, intercept = 0) +
  theme_bw()
```


## Challenge: Gradient Boosted Trees


```{r, eval = FALSE}
library(xgboost)
dtrain <- xgb.DMatrix(data = as.matrix(select(trainDF, -toxicity)), label = trainDF$toxicity)
dtest <- xgb.DMatrix(data = as.matrix(select(testDF, -toxicity)), label = testDF$toxicity)
watch <- list(train = dtrain, test = dtest)
```

## Cross Validation and Tuning

```{r, eval = FALSE}
paramList <- list(
  list(eta = 0.3, max_depth = 6),
  list(eta = 0.09, max_depth = 4),
  list(eta = 0.08, max_depth = 4),
  list(eta = 0.07, max_depth = 4),
  list(eta = 0.06, max_depth = 4),
  list(eta = 0.05, max_depth = 4),
  list(eta = 0.04, max_depth = 4),
  list(eta = 0.03, max_depth = 4),
  list(eta = 0.02, max_depth = 4),
  list(eta = 0.01, max_depth = 4),
  list(eta = 0.09, max_depth = 2),
  list(eta = 0.08, max_depth = 2),
  list(eta = 0.07, max_depth = 2),
  list(eta = 0.06, max_depth = 2),
  list(eta = 0.05, max_depth = 2),
  list(eta = 0.04, max_depth = 2),
  list(eta = 0.03, max_depth = 2),
  list(eta = 0.02, max_depth = 2),
  list(eta = 0.01, max_depth = 2),
  list(eta = 0.08, max_depth = 4, colsample_bytree = 0.8),
  list(eta = 0.08, max_depth = 4, subsample = 0.8),
  list(eta = 0.04, max_depth = 4, subsample = 0.5),
  list(eta = 0.04, max_depth = 4, subsample = 0.5, colsample_bytree = 0.8)
)
bestResults <- tibble()
for(i in seq(length(paramList))) {
  toxCV <- xgb.cv(params = paramList[[i]], 
                  data = dtrain, 
                  nrounds = 500, 
                  nfold = 5,
                  early_stopping_rounds = 10,
                  verbose = TRUE,
                  print_every_n = 10)
  bestResults <- rbind(bestResults, toxCV$evaluation_log[toxCV$best_iteration])
}
```


```{r, eval = FALSE}
set.seed(3412)
paramList <- lapply(1:50, function(i) {
  list(eta = runif(1, 0.01, 0.3),
       max_depth = sample(3:10, 1))
})
bestResults <- tibble()
for(i in seq(length(paramList))) {
  toxCV <- xgb.cv(params = c(paramList[[i]], 
                             objective = "reg:squarederror",
                             tree_method = "hist"), 
                  data = dtrain, 
                  nrounds = 500, 
                  nfold = 10,
                  early_stopping_rounds = 10,
                  verbose = TRUE,
                  print_every_n = 10)
  bestResults <- rbind(bestResults, toxCV$evaluation_log[toxCV$best_iteration])
}
```

```{r, eval = FALSE}
set.seed(3412)
paramList <- lapply(1:50, function(i) {
  list(eta = runif(1, 0.01, 0.3),
       gamma = runif(1, 0, 0.2))
})
bestResults <- tibble()
for(i in seq(length(paramList))) {
  toxCV <- xgb.cv(params = c(paramList[[i]], 
                             objective = "reg:squarederror",
                             tree_method = "hist"), 
                  data = dtrain, 
                  nrounds = 500, 
                  nfold = 10,
                  early_stopping_rounds = 10,
                  verbose = TRUE,
                  print_every_n = 10)
  bestResults <- rbind(bestResults, toxCV$evaluation_log[toxCV$best_iteration])
}
```

## First, select a value of eta

Choose values of eta and nrounds taking into account performance/time tradeoff, and
stick with this choice.

```{r, eval = FALSE}
paramList <- lapply(0:50, function(i) {
   list(eta = 0.001 + i/50 * 0.4)
})
bestResults <- tibble()
for(i in seq(length(paramList))) {
  toxCV <- xgb.cv(params = c(paramList[[i]], 
                             objective = "reg:squarederror",
                             tree_method = "hist"), 
                  data = dtrain, 
                  nrounds = 500, 
                  nfold = 10,
                  early_stopping_rounds = 10,
                  verbose = TRUE,
                  print_every_n = 10)
  bestResults <- rbind(bestResults, toxCV$evaluation_log[toxCV$best_iteration])
}
```

Let's use a learning rate of 0.1 for parameter tuning (nrounds around 56). Once we have selected the parameters, we will use a learning rate of 0.03 for training (nrounds 208..293). 

## Grid search

```{r, eval = FALSE}
paramGrid <- expand.grid(
  max_depth = 4:12,
  gamma = seq(0, 0.4, length.out = 10)
)
paramList <- lapply(1:nrow(paramGrid), function(i) {
  list(max_depth = paramGrid$max_depth[i],
       gamma = paramGrid$gamma[i])
})
bestResults <- tibble()
for(i in seq(length(paramList))) {
  cat("Testing case ", i, " out of ", length(paramList), ".\n")
  toxCV <- xgb.cv(params = c(paramList[[i]], 
                             eta = 0.1, # for parameter tuning
                             objective = "reg:squarederror",
                             tree_method = "approx"), 
                  data = dtrain, 
                  nrounds = 100, 
                  nfold = 10,
                  early_stopping_rounds = 10,
                  verbose = TRUE,
                  print_every_n = 10)
  bestResults <- rbind(bestResults, toxCV$evaluation_log[toxCV$best_iteration])
}
depth_gamma <- bind_cols(paramGrid, bestResults)
```


```{r, eval = FALSE}
paramGrid <- expand.grid(
  max_depth = 12,
  gamma = seq(0, 0.02, length.out = 10)
)
paramList <- lapply(1:nrow(paramGrid), function(i) {
  list(max_depth = paramGrid$max_depth[i],
       gamma = paramGrid$gamma[i])
})
bestResults <- tibble()
for(i in seq(length(paramList))) {
  cat("Testing case ", i, " out of ", length(paramList), ".\n")
  toxCV <- xgb.cv(params = c(paramList[[i]], 
                             eta = 0.1, # for parameter tuning
                             objective = "reg:squarederror",
                             tree_method = "approx"), 
                  data = dtrain, 
                  nrounds = 100, 
                  nfold = 10,
                  early_stopping_rounds = 10,
                  verbose = TRUE,
                  print_every_n = 10)
  bestResults <- rbind(bestResults, toxCV$evaluation_log[toxCV$best_iteration])
}
depth_gamma2 <- bind_cols(paramGrid, bestResults)
```

Best: 
gamma = 0, depth = 5, 4, 6
gamma = .004, depth = 12?

Next: try adding some randomness?

```{r, eval = FALSE}
set.seed(3444)
paramGrid <- expand.grid(
  subsample = seq(0.5, 1, by = 0.1),
  colsample_bytree = seq(0.5, 1, by = 0.1)
)
paramList <- lapply(1:nrow(paramGrid), function(i) {
  list(subsample = paramGrid$subsample[i],
       colsample_bytree = paramGrid$colsample_bytree[i])
})
bestResults <- tibble()
for(i in seq(length(paramList))) {
  cat("Testing case", i, "out of", length(paramList), "\n")
  toxCV <- xgb.cv(params = c(paramList[[i]], 
                             eta = 0.1, # for parameter tuning
                             gamma = 0,
                             max_depth = 5,
                             objective = "reg:squarederror",
                             tree_method = "approx"), 
                  data = dtrain, 
                  nrounds = 150, 
                  nfold = 10,
                  early_stopping_rounds = 10,
                  verbose = TRUE,
                  print_every_n = 10)
  bestResults <- rbind(bestResults, toxCV$evaluation_log[toxCV$best_iteration])
}
randomness <- bind_cols(paramGrid, bestResults)
```

# max depth and min child weight

```{r, eval = FALSE}
source("loadtopodata.R")
paramGrid <- expand.grid(
  max_depth = 4:12,
  min_child_weight = c(1, seq(35, 80, by = 5))
)
paramList <- lapply(split(paramGrid, 1:nrow(paramGrid)), as.list)
bestResults <- tibble()
for(i in seq(length(paramList))) {
  cat("Testing case ", i, " out of ", length(paramList), ".\n")
  toxCV <- xgb.cv(params = c(paramList[[i]], 
                             eta = 0.1, # for parameter tuning
                             objective = "reg:squarederror",
                             tree_method = "hist"), 
                  data = dtrain,
                  nrounds = 100, 
                  nfold = 10,
                  early_stopping_rounds = 10,
                  verbose = TRUE,
                  print_every_n = 10)
  gc()
  bestResults <- rbind(bestResults, toxCV$evaluation_log[toxCV$best_iteration])
}
depth_childweight <- bind_cols(paramGrid, bestResults)
```

## Final Training and Testing

Let's go with max_depth 5 and min_child_weight 65. (or 8,65, or 7,65, or 10,80)

```{r, eval = FALSE}
source("loadtopodata.R")
gbm <- xgb.train(data = dtrain, watchlist = watch, verbose = 0,
               nrounds = 500,
               early_stopping_rounds = 25,
               max_depth = 8,
               min_child_weight = 65,
               eta = 0.03
               )
gbm$evaluation_log %>% 
  pivot_longer(cols = c(train_rmse, test_rmse), names_to = "RMSE") %>% 
  ggplot(aes(x = iter, y = value, color = RMSE)) + geom_line()
tail(gbm$evaluation_log)
```




```{r, eval = FALSE}
pTox <- predict(gbm, as.matrix(select(testDF, -toxicity)))
gbErrors <- pTox - testDF$toxicity
tibble(`Predicted Toxicity` = pTox, Error = gbErrors) %>%
  ggplot(aes(x = `Predicted Toxicity`, y = Error))  +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 0, intercept = 0) +
  theme_bw()
```

```{r, eval = FALSE}
importance(toxfor) %>% 
  as_tibble(rownames = "Variable") %>% 
  arrange(desc(IncNodePurity))
```

```{r, eval = FALSE}
xgb.importance(model = gbm)
```


# delete ^^^^




# Hyperparameters


# Cross Validation


# Red Wine Data, revisited

## Reload Red Wine Training/Test set

```{r, message=FALSE, results=FALSE, warning=FALSE}
library(tidyverse)
library(here)
library(xgboost)
wine <- read_csv(here("data", "wine.csv"))
redwine <- wine %>% dplyr::slice(1:1599) 
trainSize <- round(0.80 * nrow(redwine))
set.seed(1234) 
trainIndex <- sample(nrow(redwine), trainSize)
trainDF <- redwine %>% dplyr::slice(trainIndex)
testDF <- redwine %>% dplyr::slice(-trainIndex)
dtrain <- xgb.DMatrix(data = as.matrix(select(trainDF, -quality)), label = trainDF$quality)
dtest <- xgb.DMatrix(data = as.matrix(select(testDF, -quality)), label = testDF$quality)
watch <- list(train = dtrain, test = dtest)
```




## Further tuning on white wine data

```{r, eval = FALSE}
source("loadwinedata.R")
paramGrid <- tibble(eta = seq(0.001, 0.4, length.out = 50))
paramList <- lapply(split(paramGrid, 1:nrow(paramGrid)), as.list)
bestResults <- tibble()
for(i in seq(length(paramList))) {
  toxCV <- xgb.cv(params = c(paramList[[i]], 
                             objective = "reg:squarederror",
                             tree_method = "hist"), 
                  data = dtrain, 
                  nrounds = 500, 
                  nfold = 10,
                  early_stopping_rounds = 10,
                  verbose = TRUE,
                  print_every_n = 10)
  gc()
  bestResults <- rbind(bestResults, toxCV$evaluation_log[toxCV$best_iteration])
}
etasearch <- bind_cols(paramGrid, bestResults)
```

All values of eta are overfitting. Let's go with 0.09 for tuning experiments. See if we can beat a test RMSE of 0.6135.

```{r, eval = FALSE}
source("loadwinedata.R")
paramGrid <- expand.grid(
  max_depth = 5:13,
  min_child_weight = seq(0, 20, by = 2)
)
paramList <- lapply(split(paramGrid, 1:nrow(paramGrid)), as.list)
bestResults <- tibble()
print(Sys.time())
for(i in seq(length(paramList))) {
  cat("   #", i, "of", length(paramList))
  gbmod <- xgb.cv(params = c(paramList[[i]], 
                             eta = 0.1, # for parameter tuning
                             objective = "reg:squarederror",
                           #  tree_method = "hist"), 
                             tree_method = "exact"), 
                  data = dtrain, 
                  nrounds = 500, 
                  nfold = 10,
                  early_stopping_rounds = 10,
                  verbose = FALSE,
                  print_every_n = 10)
  bestResults <- rbind(bestResults, gbmod$evaluation_log[gbmod$best_iteration])
  gc() # Free unused memory after each loop iteration
}
cat("Finished.\n")
print(Sys.time())
depth_childweight <- bind_cols(paramGrid, bestResults)
```

Looks like max_depth = 3 and min_child_weight = 50 doesn't overfit too bad. Try it:

However max_depth = 12 and min_child_weight = 8 gives a test rmse of 0.6058.


```{r, eval = FALSE}
gbmod <- xgb.train(data = dtrain, watchlist = watch, verbose = 0,
               nrounds = 1000,
               early_stopping_rounds = 25,
               max_depth = 12,
               min_child_weight = 8,
               eta = 0.05
               )
gbmod$evaluation_log %>% 
  pivot_longer(cols = c(train_rmse, test_rmse), names_to = "RMSE") %>% 
  ggplot(aes(x = iter, y = value, color = RMSE)) + geom_line()
tail(gbmod$evaluation_log)
```
Best so far: test_rmse = 0.629. 


```{r, eval = FALSE}
pww <- predict(gbmod, as.matrix(select(testDF, -quality)))
gbErrors <- pww - testDF$quality
tibble(`Predicted Quality` = pww, Error = gbErrors) %>%
  ggplot(aes(x = `Predicted Quality`, y = Error))  +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 0, intercept = 0) +
  theme_bw()
```


```{r, eval = FALSE}
xgb.importance(model = gbmod)
```

## Tune using max_depth only?

```{r, eval = FALSE}
source("loadwinedata.R")
paramGrid <- expand.grid(
  max_depth = 1:20
)
paramList <- lapply(split(paramGrid, 1:nrow(paramGrid)), as.list)
bestResults <- tibble()
for(i in seq(length(paramList))) {
  cat("   #", i, "of", length(paramList))
  gbmod <- xgb.cv(params = c(paramList[[i]], 
                             eta = 0.1, # for parameter tuning
                             objective = "reg:squarederror",
                             tree_method = "hist"), 
                  data = dtrain, 
                  nrounds = 500, 
                  nfold = 10,
                  early_stopping_rounds = 10,
                  verbose = FALSE,
                  print_every_n = 10)
  gc()
  bestResults <- rbind(bestResults, gbmod$evaluation_log[gbmod$best_iteration])
}
cat("Finished.\n")
depth_only <- bind_cols(paramGrid, bestResults)
```

# Lauree++ recommended method

https://sites.google.com/view/lauraepp/parameters

## Tune using max_depth and max_leaves

```{r, eval = FALSE}
source("loadwinedata.R")
paramGrid <- expand.grid(
  max_depth = seq(10, 30, by = 2),
  max_leaves = c(15, 31, 63, 127, 255, 511, 1023, 2047, 4095)
)
paramList <- lapply(split(paramGrid, 1:nrow(paramGrid)), as.list)
bestResults <- tibble()
for(i in seq(length(paramList))) {
  cat("   #", i, "of", length(paramList))
  gbmod <- xgb.cv(params = c(paramList[[i]], 
                             eta = 0.1, # for parameter tuning
                             objective = "reg:squarederror",
                             tree_method = "hist"), 
                  data = dtrain, 
                  nrounds = 500, 
                  nfold = 10,
                  early_stopping_rounds = 10,
                  verbose = FALSE,
                  print_every_n = 10)
  gc()
  bestResults <- rbind(bestResults, gbmod$evaluation_log[gbmod$best_iteration])
}
cat("Finished.\n")
depth_leaves <- bind_cols(paramGrid, bestResults)
```

Best test-rmse is 12,63, followed by 26,255, 28,255.

## Grid for random sampling.

```{r, eval = FALSE}
source("loadwinedata.R")
paramGrid <- expand.grid(
  subsample = seq(0.5, 1, by = 0.1),
  colsample_bytree = seq(0.5, 1, by = 0.1)
)
paramList <- lapply(split(paramGrid, 1:nrow(paramGrid)), as.list)
bestResults <- tibble()
set.seed(2341)
for(i in seq(length(paramList))) {
  cat("   #", i, "of", length(paramList))
  gbmod <- xgb.cv(params = c(paramList[[i]], 
                             max_depth = 28, 
                             max_leaves = 255,
                             eta = 0.1, # for parameter tuning
                             objective = "reg:squarederror",
                             tree_method = "hist"), 
                  data = dtrain, 
                  nrounds = 500, 
                  nfold = 10,
                  early_stopping_rounds = 10,
                  verbose = FALSE,
                  print_every_n = 10)
  gc()
  bestResults <- rbind(bestResults, gbmod$evaluation_log[gbmod$best_iteration])
}
cat("Finished.\n")
randrowcol <- bind_cols(paramGrid, bestResults)
```

0.602 (.9,.6)
.603  (.8, .9)

## Try tuning gamma?

```{r, eval = FALSE}
source("loadwinedata.R")
paramGrid <- expand.grid(
  gamma = seq(0, 0.1, by = 0.01)
)
paramList <- lapply(split(paramGrid, 1:nrow(paramGrid)), as.list)
bestResults <- tibble()
set.seed(2341)
for(i in seq(length(paramList))) {
  cat("   #", i, "of", length(paramList))
  gbmod <- xgb.cv(params = c(paramList[[i]], 
                             max_depth = 28, 
                             max_leaves = 255,
                             subsample = 0.9,
                             colsample_bytree = 0.6,
                             eta = 0.1, # for parameter tuning
                             objective = "reg:squarederror",
                             tree_method = "hist"), 
                  data = dtrain, 
                  nrounds = 500, 
                  nfold = 10,
                  early_stopping_rounds = 10,
                  verbose = FALSE,
                  print_every_n = 10)
  gc()
  bestResults <- rbind(bestResults, gbmod$evaluation_log[gbmod$best_iteration])
}
cat("Finished.\n")
gammatest <- bind_cols(paramGrid, bestResults)
```

gamma = 0.06 gives a little bit of improvement.

## Final check against test set

```{r, eval = FALSE}
gbmod <- xgb.train(data = dtrain, watchlist = watch, verbose = 0,
               nrounds = 10000,
               early_stopping_rounds = 50,
               max_depth = 28,
               max_leaves = 255,
               subsample = 0.9,
               colsample_bytree = 0.6,
               gamma = 0.06,
               eta = 0.001
               )
gbmod$evaluation_log %>% 
  pivot_longer(cols = c(train_rmse, test_rmse), names_to = "RMSE") %>% 
  ggplot(aes(x = iter, y = value, color = RMSE)) + geom_line()
tail(gbmod$evaluation_log)
```
Best so far: test_rmse = 0.614885. 


```{r, eval = FALSE}
pww <- predict(gbmod, as.matrix(select(testDF, -quality)))
gbErrors <- pww - testDF$quality
tibble(`Predicted Quality` = pww, Error = gbErrors) %>%
  ggplot(aes(x = `Predicted Quality`, y = Error))  +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 0, intercept = 0) +
  theme_bw()
```





## Facebook data: overfitting ok?

Data from UCI: https://archive.ics.uci.edu/ml/machine-learning-databases/00363/

Cambridge Spark tutorial: https://blog.cambridgespark.com/hyperparameter-tuning-in-xgboost-4ff9100a3b2f

```{r eval=FALSE}
library(tidyverse)
library(here)
facebook <- read_csv(here("data", "facebook_comments", "Features_Variant_1.csv"), 
                     col_names = FALSE)
trainSize <- round(0.90 * nrow(facebook))
set.seed(123) 
trainIndex <- sample(nrow(facebook), trainSize)
trainDF <- facebook %>% dplyr::slice(trainIndex)
testDF <- facebook %>% dplyr::slice(-trainIndex)
  
library(xgboost)
dtrain <- xgb.DMatrix(data = as.matrix(select(trainDF, -X54)), label = trainDF$X54)
dtest <- xgb.DMatrix(data = as.matrix(select(testDF, -X54)), label = testDF$X54)
watch <- list(train = dtrain, test = dtest)

mean(abs(testDF$X54 - mean(trainDF$X54))) # should be 11.31?
```

```{r eval=FALSE}
fmod <- xgb.train(data = dtrain, watchlist = watch, verbose = 0,
               nrounds = 1000,
               early_stopping_rounds = 10,
               max_depth = 6,
               min_child_weight = 1,
               eta = 0.3,
               subsample = 1,
               colsample_bytree = 1,
               objective = "reg:squarederror", # reg:linear is deprecated
               eval_metric = "mae"
               )
print(fmod) # MAE is a little different (4.31 in tutorial)
```

```{r eval=FALSE}
fmod$evaluation_log %>% 
  pivot_longer(cols = c(train_mae, test_mae), names_to = "MAE") %>% 
  ggplot(aes(x = iter, y = value, color = MAE)) + geom_line()
tail(fmod$evaluation_log)
```
Indeed there is quite a lot of overfitting.

```{r, eval = FALSE}
predX54 <- predict(fmod, as.matrix(select(testDF, -X54)))
gbErrors <- predX54 - testDF$X54
tibble(Predicted = predX54, Error = gbErrors) %>%
  ggplot(aes(x = Predicted, y = Error))  +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 0, intercept = 0) +
  theme_bw()
```

Here's the "tuned" model:

```{r, eval = FALSE}
fmod <- xgb.train(data = dtrain, watchlist = watch, verbose = 0,
               nrounds = 1000,
               early_stopping_rounds = 10,
               max_depth = 10,
               min_child_weight = 6,
               eta = 0.01,
               subsample = 0.8,
               colsample_bytree = 1,
               objective = "reg:squarederror", # reg:linear is deprecated
               eval_metric = "mae"
               )
print(fmod) # MAE is a little different (3.90 in tutorial)
```

```{r eval=FALSE}
fmod$evaluation_log %>% 
  pivot_longer(cols = c(train_mae, test_mae), names_to = "MAE") %>% 
  ggplot(aes(x = iter, y = value, color = MAE)) + geom_line()
tail(fmod$evaluation_log)
```

Overfitting has improved somewhat, but it's still there. 